---
title: "Exercise-11"
format: html
editor: visual
---

## Practice Model Selection

### Step 1. Loading in data set and replacing all values (-999) with NA

```{r}

library(tidyverse)
library(broom)
library(MuMIn)
library(skimr)

f <- "https://raw.githubusercontent.com/difiore/ada-datasets/main/Mammal_lifehistories_v2.txt"
d <- read_tsv(f, col_names = TRUE)

head(d)

skim(d)
```

##### after some exploratory analysis using the skimr () package we see that family ,species, and genus are categorical and the rest are numeric variables . 

#### replacing the -999 values with NA 

```{r}

library(naniar) #trying the naninar package with the replace_wiiht_na_all function

d <- d %>% replace_with_na_all(condition = ~.x == -999)

head(d) # we see that it worked ! 
```

### Step 2. Drop the Variables litter size and refs

```{r}

d <- d |> select(-c("litter size", "refs"))

d # it has been dropped
```

### Step 3. log transform all of the other numeric variables.

```{r}
library(dplyr)
d <- d |> mutate(across(where(is.numeric), log))
d # checking work yes we did log transfolrm all of the numeric variables
```
### Step 4. Regress the (now log transformed) age [gestation(mo), weaning(mo), AFR(mo) (i.e., age at first reproduction), and max. life(mo) (i.e., maximum lifespan)] and mass [newborn(g) and wean mass(g)] variables on (now log transformed) overall body mass(g)

